Type Refinements for the Working Class

### 2015-11-20 paper.tex
fix up a missing label
Define applicative approximation (#3)
@jozefg

%% fonts
\usepackage{pxfonts}
\usepackage{eulervm}
\usepackage[
  activate={true,nocompatibility},
  kerning=true,
  spacing=true,
  tracking=true
]{microtype}
\DeclareSymbolFont{bbold}{U}{bbold}{m}{n}
\DeclareSymbolFontAlphabet{\mathbbold}{bbold}

@@

\newcommand\MkValence[3]{%
  \ifthenelse{\equal{#1}{}}{}{%
    \{#1\}%
  }%
  \ifthenelse{\equal{#2}{}}{}{%
    [#2]%
  }.\, #3%
}
\newcommand\MkBTm[3]{\{#1\}[#2].\, #3}
\newcommand\MkArity[2]{(#1)\, #2}
\newcommand\IsArity[1]{\IMode{#1}\ \textit{arity}}
\newcommand\IsValence[1]{\IMode{#1}\ \textit{valence}}
\newcommand\IsSort[1]{\IMode{#1}\ \textit{sort}}
\newcommand\Leq[2]{\IMode{#1}\leq\IMode{#2}}

@@

\newcommand\Rename[2]{%
  #2\upharpoonright#1%
}

\newcommand\IsRenaming[3]{%
  \Rename{\IMode{#1}}{\IMode{#2}}%
  \leadsto\OMode{#3}%
}

@@

\newcommand\EvalN[5]{\IMode{#1}\parallel\IMode{#4}\Downarrow^{\OMode{#3}}_{\IMode{#2}}\OMode{#5}}
\newcommand\Eval[4]{\EvalN{#1}{#2}{}{#3}{#4}}
\newcommand\Exprs{\mathcal{E}}
\newcommand\Values{\mathcal{V}}
\newcommand\Naturals{\mathbb{N}}
\newcommand\NaturalsPsh{\mathcal{N}}

@@

\section{The Ambient Logical Framework}

@@

In this section we generalize Howe's notion of lazy computation
system~\cite{howe:1989} to the multi-sorted, symbol-parameterized setting. An
\emph{lazy computation language} is an abt signature
$\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$ along with a distinguished copresheaf
$\Of{\CanOperators}{\Sets^{\SCtx\times\Discrete{\Arities}}}$ of
\emph{canonical} operators such that $\IsSubsetEq{\CanOperators}{\Operators}$.
For a lazy computation language
$\Match{L}{\Tuple{\Sorts,\Operators,\CanOperators}}$, let us define the
presheaves of closed expressions and closed values as follows:
\begin{align*}
  \ADefine{\Exprs_\tau(\Upsilon)}{
    \MkSet{M \mid \IsAbtUnmoded{\cdot}{\Upsilon}{\cdot}{M}{\tau}}
  }\\
  \ADefine{\Values_\tau(\Upsilon)}{
    \MkSet{
      M\equiv\App{\vartheta}{\vec{E}}
      \mid M\in\Exprs_\tau(\Upsilon)\land \exists a. \CanOperators\Pair{\Upsilon}{a}\ni\vartheta
    }
  }
\end{align*}

\begin{remark}
  If read as a judgment, $\Exprs_\tau$ (resp.\ $\Values_\tau$) means ``there is
  a closed expression (resp.\ value) of sort $\tau$'', and has as its evidence any closed
  expression (resp.\ value) of sort $\tau$.
\end{remark}

Then, a \emph{lazy computation system} (lcs) is a lazy computation language
$\Match{L}{\Pair{\Sigma}{\CanOperators}}$ along with an $\SCtx$-indexed
$\equiv_\alpha$-functional evaluation relation
$\EvalN{\Upsilon}{\tau}{n}{M}{N}$ presupposing $\Member{n}{\Naturals}$,
$\Member{M}{\Exprs_\tau(\Upsilon)}$ and $\Member{N}{\Values_\tau(\Upsilon)}$,
expressing that $M$ evaluates to $N$ in $n$ steps.  We will write
$\Eval{\Upsilon}{\tau}{M}{N}$ to mean that there exists an $n$ such that
$\EvalN{\Upsilon}{\tau}{n}{M}{N}$.

@@

A \emph{global internal (family of) judgments} is a section
$\Of{J}{\int_\SCtx {\Omega^{\mathcal{J}}}}\cong
\IMode{\int_\Upsilon\Hom{\mathbb{J}}{\Yoneda{\Upsilon}\times\mathcal{J}}{\Omega}}$,
which can be externalized as a ``total'' judgment $\Of{\Embed{J}}{\mathbb{J}}$
for whose derivations at $\Upsilon$ are the arrows
$\Of{\phi}{\Hom{\mathbb{J}}{\Yoneda{\Upsilon}}{\mathbbold{1}\times\mathcal{J}}}$
such that the following diagram commutes:
\[
  \begin{tikzcd}[cramped,sep=huge]
    \IMode{\Yoneda{\Upsilon}}
      \arrow[r, dashed, "\OMode{\phi}"]
      \arrow[rd, "\IMode{J_\Upsilon\Pair{1}{\pi_2\circ\phi}}"']
    & \IMode{\mathbbold{1}\times\mathcal{J}} \arrow[d, "\IMode{\mathbf{true}\circ\pi_1}"] \\
    & \IMode{\Omega}
  \end{tikzcd}
\]

>>>>

%% fonts
\usepackage{pxfonts}
\usepackage{eulervm}
\usepackage[
  activate={true,nocompatibility},
  kerning=true,
  spacing=true,
  tracking=true
]{microtype}
\usepackage{bbold}

@@

\newcommand\MkValence[3]{%
  \ifthenelse{\equal{#1}{}}{}{%
    \{#1\}%
  }%
  \ifthenelse{\equal{#2}{}}{}{%
    [#2]%
  }.\, #3%
}
\newcommand\MkBTm[3]{\mathbb{\lambda}\{#1\}[#2].\, #3}
\newcommand\MkArity[2]{(#1)\, #2}
\newcommand\IsArity[1]{\IMode{#1}\ \textit{arity}}
\newcommand\IsValence[1]{\IMode{#1}\ \textit{valence}}
\newcommand\IsSort[1]{\IMode{#1}\ \textit{sort}}
\newcommand\Leq[2]{\IMode{#1}\leq\IMode{#2}}

@@

\newcommand\Rename[3]{%
  \left\{#1\, /\, #2\right\}#3%
}

@@

\newcommand\EvalN[5]{\IMode{#1}\parallel\IMode{#4}\Downarrow^{\OMode{#3}}_{\IMode{#2}}\OMode{#5}}
\newcommand\Eval[4]{\EvalN{#1}{#2}{}{#3}{#4}}
\newcommand\Exprs{\mathbf{E}}
\newcommand\Values{\mathbf{V}}
\newcommand\BTms{\mathbf{B}}
\newcommand\Naturals{\mathbb{N}}
\newcommand\NaturalsPsh{\mathcal{N}}

@@

\section{The Ambient Logical Framework}
\label{section:ambient-lf}

@@

In this section we generalize Howe's notion of lazy computation
system~\cite{howe:1989} to the multi-sorted, symbol-parameterized setting. An
\emph{lazy computation language} is an abt signature
$\Match{\Sigma}{\Pair{\Sorts}{\Operators}}$ along with a distinguished
copresheaf $\Of{\CanOperators}{\Sets^{\SCtx\times\Discrete{\Arities}}}$ of
\emph{canonical} operators such that $\IsSubsetEq{\CanOperators}{\Operators}$.
For a lazy computation language
$\Match{L}{\Tuple{\Sorts,\Operators,\CanOperators}}$, let us define the
covariant presheaves on $\SCtx\times\Ctx$ of open expressions, open values, and open bound terms
as follows:
\begin{align*}
  \ADefine{\Exprs_\tau(\Upsilon\parallel\Gamma)}{
    \MkSet{M \mid \IsAbtUnmoded{\cdot}{\Upsilon}{\Gamma}{M}{\tau}}
  }\\
  \ADefine{\Values_\tau(\Upsilon\parallel\Gamma)}{
    \MkSet{
      M\equiv\App{\vartheta}{\vec{E}}
      \mid M\in\Exprs_\tau(\Upsilon\parallel\Gamma)\land \exists a. \CanOperators\Pair{\Upsilon}{a}\ni\vartheta
    }
  }\\
  \ADefine{
    \BTms_{v}(\Upsilon\parallel\Gamma)
  }{
    \MkSet{E \mid \IsAbtUnmoded{\cdot}{\Upsilon}{\Gamma}{E}{v}}
  }
\end{align*}

We'll write $\Exprs_\tau(\Upsilon)$, $\Values_\tau(\Upsilon)$ and
$\BTms_v(\Upsilon)$ for $\Exprs_\tau(\Upsilon\parallel\cdot)$,
$\Values_\tau(\Upsilon\parallel\cdot)$ and $\BTms_v(\Upsilon\parallel\cdot)$
respectively, viewed as covariant presheaves on $\SCtx$. Then, a \emph{lazy
computation system} (lcs) is a lazy computation language
$\Match{L}{\Pair{\Sigma}{\CanOperators}}$ along with an $\SCtx$-indexed
$\equiv_\alpha$-functional evaluation relation
$\EvalN{\Upsilon}{\tau}{n}{M}{N}$ presupposing $\Member{n}{\Naturals}$,
$\Member{M}{\Exprs_\tau(\Upsilon)}$ and $\Member{N}{\Values_\tau(\Upsilon)}$,
expressing that $M$ evaluates to $N$ in $n$ steps.  We will write
$\Eval{\Upsilon}{\tau}{M}{N}$ to mean that there exists an $n$ such that
$\EvalN{\Upsilon}{\tau}{n}{M}{N}$.

In any topos $\mathcal{E}$, for an object $X$ we can define the object of
relations on $X$ as the exponential $\Define{\Pow{X}}{\Omega^X}$. Then, the
data of such a relation is contained in a natural transformation
$\Of{R}{\Hom{\mathcal{E}}{\mathbb{1}}{\Pow{X}}}$.

\newcommand\SBinRel[4]{\IMode{#2}\parallel\IMode{#3}\mathrel{#1}\IMode{#4}}

Fix a sort-indexed family of binary relations on closed expressions
$\Of{R_\tau}{\Hom{\Sets^{\SCtx}}{\mathbb{1}}{\Pow{\Exprs_\tau^2}}}$; we will
also write the relation as a judgment scheme
$\SBinRel{R_\tau}{\Upsilon}{M}{N}$. Now, we can always extend $R_\tau$ to a new
family of relations
$\Of{R_v}{\Hom{\Sets^{\SCtx}}{\mathbb{1}}{\Pow{\BTms_v^2}}}$ for any valence
$\Match{v}{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}$, defined pointwise as
follows:
\[
  \infer{
    \SBinRel{R_{\MkValence{\vec{\sigma}}{\vec{\tau}}{\tau}}}{\Upsilon}{
      \MkBTm{\vec{u}}{\vec{x}}{M}
    }{
      \MkBTm{\vec{v}}{\vec{y}}{N}
    }
  }{
    \begin{array}{l}
      \forall\IMode{\vec{w}}\mathrel{\#}\IMode{\Dom{\Upsilon}\cup\vec{u}\cup\vec{v}}.\quad
      \forall\Of{\vec{X}}{\Exprs_\tau^{[\vec{\tau}]}(\Upsilon,\vec{w}:\vec{\sigma})}.\\
      \SBinRel{R_\tau}{\Upsilon,\vec{w}:\vec{\sigma}}{
        \Subst{\vec{X}}{\vec{x}}{\Rename{\vec{w}}{\vec{u}}{M}}
      }{
        \Subst{\vec{X}}{\vec{y}}{\Rename{\vec{w}}{\vec{v}}{N}}
      }
    \end{array}
  }
\]

As a convenience, we'll also define this judgment over vectors of bound terms and valences:
\[
  \infer{
    \SBinRel{R_{\vec{v}}}{\Upsilon}{\vec{E}}{\vec{F}}
  }{
    \forall\Member{(E,F,v)}{(\vec{E},\vec{F},\vec{v})}.\
    \SBinRel{R_v}{\Upsilon}{E}{F}
  }
\]

Now, we can extend $R_\tau$ to a new relation $[R_\tau]$ on closed expressions
which respects a single ``layer'' of computation, defined pointwise in the following way:
\begin{gather*}
  \infer{
    \SBinRel{[R_\tau]}{\Upsilon}{M}{N}
  }{
    \Eval{\Upsilon}{\tau}{M}{\App{\vartheta}{\Vec{E}}} &
    \Eval{\Upsilon}{\tau}{N}{\App{\vartheta}{\Vec{F}}} &
    \Lookup{\CanOperators}{\vartheta}{\MkArity{\vec{v}}{\tau}} &
    \SBinRel{R_{\vec{v}}}{\Upsilon}{\vec{E}}{\vec{F}}
  }
\end{gather*}

Because $[-_\tau]$ is monotonic, it has a greatest fixed point, which we will
call \emph{applicative approximation},
$\Of{\preceq_\tau}{\Hom{\Sets^\SCtx}{\mathbb{1}}{\Pow{\Exprs_\tau^2}}}$,
written pointwise as $\SBinRel{\preceq_\tau}{\Upsilon}{M}{N}$.

%%% Now, let us consider relations on open terms; in general, it will be easiest
%%% for us to situate our constructions on relations in the presheaf topos
%%% $\Define{\widehat{\HypCat}}{\Sets^{\SCtx\times\Ctx}}$. Then, for any two
%%% objects objects $\Of{X,Y}{\widehat{\HypCat}}$, we can form the object of
%%% relations between them as $\Define{\Pow{X\times Y}}{\Omega^{X\times Y}}$.
%%%
%% . A binary relation $R_\tau$
%% between open expressions of sort $\tau$ can be thought of as a global section
%% of the exponential object
%% $\Match{\Pow{\Exprs_\tau\times\Exprs_\tau}}{\Omega^{\Exprs_\tau\times\Exprs_\tau}}$
%% of the presheaf topos $\Define{\widehat{\HypCat}}{\Sets^{\SCtx\times\Ctx}}$

@@

A \emph{global internal (family of) judgments} is a section
$\Of{J}{\int_\SCtx {\Omega^{\mathcal{J}}}}\cong
\IMode{\int_\Upsilon\Hom{\mathbb{J}}{\Yoneda{\Upsilon}\times\mathcal{J}}{\Omega}}$,
which can be externalized as a ``total'' judgment $\Of{\Embed{J}}{\mathbb{J}}$
for whose derivations at $\Upsilon$ are the arrows
$\Of{\phi}{\Hom{\mathbb{J}}{\Yoneda{\Upsilon}}{\mathbb{1}\times\mathcal{J}}}$
such that the following diagram commutes:
\[
  \begin{tikzcd}[cramped,sep=huge]
    \IMode{\Yoneda{\Upsilon}}
      \arrow[r, dashed, "\OMode{\phi}"]
      \arrow[rd, "\IMode{J_\Upsilon\Pair{1}{\pi_2\circ\phi}}"']
    & \IMode{\mathbb{1}\times\mathcal{J}} \arrow[d, "\IMode{\mathbf{true}\circ\pi_1}"] \\
    & \IMode{\Omega}
  \end{tikzcd}
\]
